\chapter{Monte Carlo Kernel}
\label{chap:kernel_mc}

\section{Problem Statement}

After the preprocessing pipeline applies stream compaction to the invasion
probability raster, each habitat is represented as a dense vector
$\mathbf{p} \in [0,1]^{N_c}$, where $p_k$ is the probability that active
cell $k$ is reached by the lava flow.  The Monte Carlo kernel estimates the
probability that the \emph{invaded fraction} of cells strictly exceeds a
critical threshold $\theta$:

\begin{equation}
  \Pr\!\left(\frac{1}{N_c}\sum_{k=0}^{N_c-1} X_k > \theta\right),
  \qquad X_k \sim \mathrm{Bernoulli}(p_k).
\end{equation}

The estimate is obtained by running $R$ independent trials and counting how
many exceed the threshold.

% ─────────────────────────────────────────────────────────────────────────────
\section{1-D Kernel}
\label{sec:kernel1d}

\subsection{Launch Scheme}

The NDRange is one-dimensional: each work-item handles one complete trial.

\begin{center}
\begin{tabular}{ll}
  \textbf{Global NDRange} & $\lceil R / \mathtt{WG\_SIZE} \rceil \times \mathtt{WG\_SIZE}$ \\
  \textbf{Local NDRange}  & $\mathtt{WG\_SIZE} = 256$ \\
\end{tabular}
\end{center}

\subsection{Algorithm for Work-Item \texorpdfstring{$r$}{r}}

\begin{enumerate}
  \item Seed MWC64X at stream position
        $\mathtt{base\_offset} + r \cdot N_c$ with length $N_c$:
        each work-item occupies a non-overlapping segment of the RNG sequence.
  \item For each cell $k = 0, \ldots, N_c - 1$: draw
        $x \sim U(0,1)$ and compute $b_k = \mathbf{1}[x \le p_k]$.
  \item Count invaded cells: $\mathtt{invaded} = \sum_k b_k$.
  \item Compute the success bit:
        $s_r = \mathbf{1}[\mathtt{invaded} / N_c > \theta]$.
  \item Participate in the work-group tree reduction;
        thread 0 issues \texttt{atomic\_add} on the global counter.
\end{enumerate}

\subsection{Source}

\begin{lstlisting}[language=OpenCL, caption={1-D kernel --- \texttt{monte\_carlo.cl}}]
__kernel __attribute__((reqd_work_group_size(WG_SIZE, 1, 1)))
void monte_carlo_run(
    __global const float* p_vec,
    __global int*         count,
    const uint            n_cells,
    const float           threshold,
    const ulong           base_offset,
    const uint            n_runs)
{
    uint r   = get_global_id(0);
    uint lid = get_local_id(0);

    int my_result = 0;
    if (r < n_runs)
        my_result = _run_trial(p_vec, n_cells, threshold,
                               base_offset + (ulong)r * n_cells);

    __local int scratch[WG_SIZE];
    scratch[lid] = my_result;
    barrier(CLK_LOCAL_MEM_FENCE);
    _tree_reduce(scratch, lid);

    if (lid == 0)
        atomic_add(count, scratch[0]);
}
\end{lstlisting}

\subsection{\texttt{p\_vec} Reads and Cache Behaviour}

At each iteration $k$, all 256 work-items in the work-group read the
\emph{same} element \texttt{p\_vec[k]}: the value is loaded into L1 cache
once and broadcast to all threads at no additional cost.

% ─────────────────────────────────────────────────────────────────────────────
\section{2-D Kernel Attempt}
\label{sec:kernel2d}

\subsection{Motivation}

In the 1-D kernel each work-item samples $N_c$ cells sequentially.  The idea
behind the 2-D kernel is to parallelise the cell axis as well: each work-item
samples exactly \emph{one} pair $(\text{run}\; r,\; \text{cell}\; k)$,
exploiting a two-dimensional NDRange.

\begin{center}
\begin{tabular}{ll}
  \textbf{Global NDRange} &
    $\lceil R/\mathtt{WG\_R}\rceil\!\times\!\mathtt{WG\_R}\ ,\quad
     \lceil N_c/\mathtt{WG\_C}\rceil\!\times\!\mathtt{WG\_C}$ \\
  \textbf{Local NDRange}  & $(\mathtt{WG\_R},\;\mathtt{WG\_C}) = (8,\;32)$ \\
\end{tabular}
\end{center}

\subsection{Need for an Intermediate Buffer}

When $N_c > \mathtt{WG\_C}$, the cells of a single run are spread across
multiple work-groups along the $k$ axis.  For example, with $N_c = 96$ and
$\mathtt{WG\_C} = 32$:

\begin{center}
\begin{tabular}{ccc}
  \textbf{Work-group} & \textbf{Runs} & \textbf{Cells} \\
  \hline
  WG$(0,0)$ & $0\ldots7$ & $0\ldots31$   \\
  WG$(0,1)$ & $0\ldots7$ & $32\ldots63$  \\
  WG$(0,2)$ & $0\ldots7$ & $64\ldots95$  \\
\end{tabular}
\end{center}

To determine whether run~0 exceeded the threshold, the partial counts from
all three work-groups must be summed.  In OpenCL~1.2
\textbf{work-groups cannot synchronise with each other} within a single
kernel launch: \texttt{barrier} is local to a work-group and has no global
visibility.  The only available global synchronisation point is the
\emph{end of a kernel launch}.  The solution is therefore to write partial
counts to a global buffer \texttt{g\_partial} and launch a second kernel
that reduces them.

\subsection{Source — Kernel 1: \texttt{mc\_partial\_sums}}

\begin{lstlisting}[language=OpenCL, caption={2-D kernel, first launch --- \texttt{monte\_carlo\_2d.cl}}]
__kernel __attribute__((reqd_work_group_size(WG_R, WG_C, 1)))
void mc_partial_sums(
    __global const float* p_vec,
    __global int*         g_partial,   /* [n_groups_c][n_runs_padded] */
    const uint            n_cells,
    const uint            n_runs,
    const ulong           base_offset,
    const uint            n_runs_padded)
{
    uint r     = get_global_id(0);
    uint k     = get_global_id(1);
    uint lid_r = get_local_id(0);
    uint lid_c = get_local_id(1);

    mwc64x_state_t rng;
    MWC64X_SeedStreams(&rng,
        base_offset + (ulong)r * n_cells + k, 1UL);

    int my_bit = 0;
    if (r < n_runs && k < n_cells)
        my_bit = (_to_float(MWC64X_NextUint(&rng)) <= p_vec[k])
                 ? 1 : 0;

    __local int scratch[WG_R * WG_C];
    scratch[lid_r * WG_C + lid_c] = my_bit;
    barrier(CLK_LOCAL_MEM_FENCE);
    _reduce_cols(scratch, lid_r, lid_c);

    if (lid_c == 0 && r < n_runs)
        g_partial[get_group_id(1) * n_runs_padded + r]
            = scratch[lid_r * WG_C];
}
\end{lstlisting}

\subsection{Intermediate Buffer Size}

The \texttt{g\_partial} buffer uses a transposed layout
$[\,n\_\text{groups\_c}\,][\,n\_\text{runs\_padded}\,]$, where

\begin{equation}
  n\_\text{groups\_c} = \left\lceil \frac{N_c}{\mathtt{WG\_C}} \right\rceil,
  \qquad
  n\_\text{runs\_padded} =
    \left\lceil \frac{R}{\mathtt{WG\_R}} \right\rceil \times \mathtt{WG\_R}.
\end{equation}

With real project data (where $N_c$ may have  $\approx 200\,000$ active
cells after stream compaction, batch size $R = 100\,000$,
$\mathtt{WG\_C} = 32$, $\mathtt{WG\_R} = 8$):

\begin{align}
  n\_\text{groups\_c}    &= \lceil 200\,000 / 32 \rceil = 6\,250 \notag\\
  n\_\text{runs\_padded} &= \lceil 100\,000 / 8  \rceil \times 8
                          = 100\,000 \notag\\
  |\texttt{g\_partial}|  &= 6\,250 \times 100\,000 \times 4\,\text{B}
                          = \mathbf{2.5\,\text{GB.}} \notag
\end{align}

This buffer is allocated on the CPU via \texttt{np.zeros}, copied to the
GPU, and released on \emph{every batch}.  With 10 batches per habitat the
cumulative transfer exceeds 25\,GB, making the 2-D kernel orders of
magnitude slower than the 1-D kernel.

\subsection{Why Barriers Do Not Solve the Problem}

The primitive \texttt{barrier(CLK\_LOCAL\_MEM\_FENCE)} synchronises only the
work-items \emph{within the same work-group}.  Work-groups \texttt{WG(0,0)},
\texttt{WG(0,1)}, and \texttt{WG(0,2)} from the example above execute
independently: OpenCL~1.2 provides no global barrier primitive that could
force them to wait for one another within a single kernel launch.  The only
global synchronisation available is the kernel launch boundary itself, which
is why the two-kernel design is necessary.

\subsection{Comparison with the 1-D Kernel}

The 2-D kernel offers no architectural advantage over the 1-D kernel for
two reasons:

\begin{enumerate}
  \item \textbf{Cache broadcast.}  In the 1-D kernel, at iteration $k$ all
        256 work-items read \texttt{p\_vec[k]}: a single cache line is
        loaded once and the value is broadcast to all threads at no extra
        cost.  The 2-D kernel reads 32 consecutive floats per cycle
        (coalesced), which is not meaningfully better in terms of effective
        throughput.

  \item \textbf{Intermediate buffer cost.}  Parallelising the cell axis
        introduces an $O(N_c \times R)$ intermediate buffer that eliminates
        any computational advantage on real-world data.
\end{enumerate}

The 1-D kernel therefore remains the optimal choice for this problem.

\subsection{Experimental Results}
\label{sec:bench}

\subsubsection*{Test Environment}

\begin{center}
\begin{tabular}{ll}
  \toprule
  \textbf{Component} & \textbf{Details} \\
  \midrule
  CPU             & Intel Core i7-12700H (20 threads, 4.70\,GHz) \\
  GPU             & NVIDIA GeForce RTX 4050 Max-Q / Mobile \\
  RAM             & 15.24\,GiB DDR5 \\
  OpenCL          & 1.2 \\
  Python          & 3.14.4 \\
  \midrule
  \texttt{MC\_RUNS}      & 1\,000 \\
  \texttt{MC\_THRESHOLD} & 0.009 \\
  \texttt{MC\_BATCHES}   & 10 \\
  Habitats processed     & 4 \\
  Kernel launches        & 40 \ ($4\ \text{habitats} \times 10\ \text{batches}$) \\
  \bottomrule
\end{tabular}
\end{center}

\subsubsection*{Timing and Bandwidth}

\begin{center}
\begin{tabular}{lrrrr}
  \toprule
  \textbf{Kernel} & \textbf{Launches} &
  \textbf{Mean (ms)} & \textbf{Min (ms)} & \textbf{Bandwidth (GB/s)} \\
  \midrule
  \texttt{monte\_carlo\_run} (1-D)    & 40 &   2.273 &  0.297 & 15.19 \\
  \midrule
  \texttt{mc\_partial\_sums} (2-D)    & 40 & 185.379 & 13.259 &  0.19 \\
  \texttt{mc\_threshold\_count} (2-D) & 40 &   0.087 &  0.010 & 12.90 \\
  \bottomrule
\end{tabular}
\end{center}

The 2-D pipeline is $\approx\!82\times$ slower than the 1-D kernel (speedup
$= 0.01\times$).  The bottleneck is \texttt{mc\_partial\_sums}, whose low
effective bandwidth (0.19\,GB/s versus 15.19\,GB/s for the 1-D kernel)
reflects the allocation and transfer latency of the 2.5\,GB
\texttt{g\_partial} buffer per batch, as discussed in
Section~\ref{sec:kernel2d}.
